The Right Way to Search Evolving Graphs

### 2016-01-29 paper.tex
use 10-point size
More edits

\documentclass[11pt,conference,compsocconf]{IEEEtran}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bm}
\usepackage[linesnumbered]{algorithm2e}
\makeatletter
\renewcommand{\@algocf@capt@plain}{above}
\makeatother
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{arrows}
\usetikzlibrary{snakes}
\usetikzlibrary{calc}
\let\emptyset\varnothing
\definecolor{hotpink}{rgb}{0.9,0,0.5}
\definecolor{deepgray}{gray}{0.35}
\definecolor{deepgray2}{gray}{0.25}
\definecolor{deepgray3}{gray}{0.10}
\definecolor{lightgray}{gray}{0.95}
\definecolor{lightgray2}{gray}{0.85}
\definecolor{purple1}{RGB}{239,229,244}
\definecolor{purple2}{RGB}{216,191,216}
\definecolor{lightblue}{rgb}{0.73,0.33,0.83}
\definecolor{lightpurple}{rgb}{.8,.2,.8}
\definecolor{textcolor}{rgb}{0,0,5}
\definecolor{blue1}{RGB}{187,217,238}
\definecolor{blue2}{RGB}{235,244,250}
\definecolor{yellow1}{RGB}{255,255,102}
\definecolor{blue3}{RGB}{63,40,96}
\definecolor{red1}{RGB}{255, 102, 102}
\definecolor{green1}{RGB}{102, 255, 102}
\newtheorem{lemma}{Lemma}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\theoremstyle{definition}
\def\proof{\par{\bf Proof}. \ignorespaces}
\def\qedsymbol{\vbox{\hrule\hbox{%
                     \vrule height1.3ex\hskip0.8ex\vrule}\hrule}}
\def\endproof{\qquad\qedsymbol\medskip\par}

@@

\begin{abstract}
Evolving graphs arise in many different contexts where the 
interrelations between data elements change over time. 
We present a breadth first search (BFS) algorithm for evolving graphs
that can track (active) nodes correctly. 
Using simple examples, we show na\"ive matrix-matrix multiplication on time-dependent adjacency matrices miscounts the number of temporal paths. 
By mapping an evolving graph to an adjacency matrix of the equivalent static graph, 
we  prove the properties of the BFS algorithm using the properties of the 
adjacency matrix.
We show an application of our algorithm in mining citation networks.

@@

We can analyze the spread of information between the players using graph
theory. In this process, the time ordering of events matters, and hence its
graph representation $G(t) = (V(t), E(t))$ must be time dependent.
Such a graph is called an ``evolving graph''~\cite{fkp89,bkmu12},
``evolving network''~ \cite{bfgm08} or
``temporal graph''~\cite{tmml09}.

Treatments of evolving graphs vary in their generality and focus.
Kivel\"a \textit{et al.}~\cite{kabg14} treats time dependence as a special
case of families of graphs with multiple interrelationships.
Others like Flajolet \textit{et al.}~\cite{fkp89} use time to index the family
of related graphs, but are not concerned with explicit time-dependent processes.
Yet others focus on incremental updates to large graphs~\cite{bkmu12}.
Here, we describe evolving graphs as a time-ordered sequence of graphs, similar
to the study of metrized graphs by Tang and coworkers~\cite{ntmm13,tmml09,tsmm09,tmml10} and of
the dynamics of communities by Grindrod, Higham and coworkers~\cite{gphe11,grihig13}.

The game described in the beginning can be encoded in an evolving graph.
The spread of information to the winner can be described in terms of traversing
this graph using discrete paths that step in both space and time.
Traversals of an ordinary (static) graph may be computed using well known methods
such as the breadth-first search (BFS). An informal description of BFS generalized
to evolving graphs can be found in \cite{tmml09}.
However, it turns out that na\"ive extensions can lead to incorrect descriptions of
the resulting graph traversals. A proper treatment requires the notion of
\emph{node activeness} to describe the set of paths that can only traverse time
or edges. We term such paths \textit{temporal paths}.
As a result, our treatment can be applied to any evolving graph, even those that
are highly dynamic with arbitrary changes to the nodes and edges.

In this paper, we explain how the BFS algorithm can be applied correctly on
evolving graphs. We also provide a counterexample demonstrating that some na\"ive
analyses in terms of discrete path sums over the time-dependent adjacency
matrices lead to incorrect results. The correct generalization to evolving graphs
is formally equivalent to BFS over a particular static graph generated by connecting
together active notes of the evolving graph. We also show that BFS over evolving
graphs can be interpreted algebraically as computing a sequence of vectors formed
by repeatedly multiplying a starting vector with the adjacency matrix of the
equivalent static graph, and in so doing reveals interesting connections between
properties of the BFS algorithm and properties of the adjacency matrix. Finally,
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving graph with 3 time stamps $t_1$, $t_2$ and $t_3$.
At each time stamp, the evolving graph is represented as a graph.
The green filled circles represent active nodes while the red circles represent
inactive nodes.}
\label{fig:eg_shortest_path}
\end{figure}

@@

  \foreach \from/\to in {n1/n4}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=70, in=40] (n11);
  \draw[dotted,->](n10) -- (n7);
   \draw[dotted,->](n11) -- (n12);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{Two temporal paths from $(1, t_1)$ to $(3,t_3)$ on the evolving graph
shown in Figure~\ref{fig:eg_shortest_path}, shown in black dashed lines.
The paths traverse only active nodes along edges, and are allowed to advance
between the same node if it is active at different times.}
\label{fig:active}
\end{figure}

@@

The example presented above in Section~\ref{sec:temporal-paths}
demonstrates how the notion of active nodes restrict the set of temporal paths
that need to be considered when traversing an evolving graph.

We now give a general description of the BFS algorithm over evolving graphs,
both directed and undirected, which correctly takes into account the structure
of temporal paths.
Our notation generalizes the standard notations for static graphs presented in
\cite{even12,kegi11}.

@@

\begin{defn}
The distance from a temporal node $(v, t)$ to a temporal node $(w, s)$ is
the $k$ for which $(w, s)$ is a $k$-forward neighbor of $(v, t)$.
\end{defn}

@@

\IEEEproof
Define the set of temporal nodes
$\tilde V_L^{[t]} = \{(v_1, t) | (v_1, v_2) \in E^{[t]}\}$, which
consists of the active nodes at time $t$ which participate on the left side of an edge.
Similarly, $\tilde V_R^{[t]} = \{ (v_2, t) | (v_1, v_2) \in E^{[t]}\}$
are the corresponding active nodes on the right side of an edge. Then
$\tilde V^{[t]} = \tilde V_L^{[t]} \cup \tilde V_R^{[t]}$ is the set of active
nodes at time $t$, and $V = \bigcup_t \tilde V^{[t]}$ is the set of all
active nodes in $G_n$.

@@

The algebraic representation of evolving graphs presented in Section \ref{sec:temp-paths-adjac}
allows us to exploit a graphical interpretation of
matrix--vector products involving the adjacency matrix~\cite{kegi11}.
If $A$ is the adjacency matrix of a (static) graph $G$ and $e_k$ is the $k$th
elementary unit vector, then the nonzero entries of $A^T e_k$ have indices that
are neighbors of $k$.
The algebraic formulation of BFS on evolving graphs follows similarly, but
requires a new kind of matrix--vector product, $\odot$, defined by
\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ can then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas all other nonzero entries represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

Consider two iterations of the BFS on $G_n$ with root $(k, t_1)$. The second
iteration requires computing the sequences

@@

The blocked matrix structure of the adjacency matrices presented here provide
interesting relationships between their matrix properties and the algorithmic
properties of BFS, made possible because of the reformulation of BFS as
repeated power iterations of the adjacency matrix in Algorithm~\ref{alg:bfsla}.
Note, however, that these matrices never need to be instantiated for practical
computations. Rather, since Algorithm~\ref{alg:bfsla} only requires the m
matrix--vector product involving the adjacency matrix, the formulation of
Algorithm~\ref{alg:bfs} provides an efficient way to exploit the block structure
of $\bm A_n$. The $\odot$ operation provides an efficient way to compute the
action of the off-diagonal products. Representing the diagonal blocks
$A^{[t]}$ as sparse matrices product further reduces the cost of BFS by exploiting
latent sparsity in graphs that show up in practical applications.

\subsection{The algebraic formulation of BFS on evolving graphs}

>>>>

\documentclass[10pt,conference,compsocconf]{IEEEtran}
\usepackage{tikz}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bm}
\usepackage[linesnumbered]{algorithm2e}
\makeatletter
\renewcommand{\@algocf@capt@plain}{above}
\makeatother
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{arrows}
\usetikzlibrary{snakes}
\usetikzlibrary{calc}
\let\emptyset\varnothing
\definecolor{hotpink}{rgb}{0.9,0,0.5}
\definecolor{deepgray}{gray}{0.35}
\definecolor{deepgray2}{gray}{0.25}
\definecolor{deepgray3}{gray}{0.10}
\definecolor{lightgray}{gray}{0.95}
\definecolor{lightgray2}{gray}{0.85}
\definecolor{purple1}{RGB}{239,229,244}
\definecolor{purple2}{RGB}{216,191,216}
\definecolor{lightblue}{rgb}{0.73,0.33,0.83}
\definecolor{lightpurple}{rgb}{.8,.2,.8}
\definecolor{textcolor}{rgb}{0,0,5}
\definecolor{blue1}{RGB}{187,217,238}
\definecolor{blue2}{RGB}{235,244,250}
\definecolor{yellow1}{RGB}{255,255,102}
\definecolor{blue3}{RGB}{63,40,96}
\definecolor{red1}{RGB}{255, 102, 102}
\definecolor{green1}{RGB}{102, 255, 102}
\newtheorem{lemma}{Lemma}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\theoremstyle{definition}
\def\proof{\par{\bf Proof}. \ignorespaces}
\def\qedsymbol{\vbox{\hrule\hbox{%
                     \vrule height1.3ex\hskip0.8ex\vrule}\hrule}}
\def\endproof{\qquad\qedsymbol\medskip\par}

@@

\begin{abstract}
Evolving graphs arise in many different contexts where the
interrelations between data elements change over time.
We present a breadth first search (BFS) algorithm for evolving graphs
that can track (active) nodes correctly.
Using simple examples, we show na\"ive matrix-matrix multiplication on
time-dependent adjacency matrices miscounts the number of temporal paths.
By mapping an evolving graph to an adjacency matrix of the equivalent static graph,
we prove the properties of the BFS algorithm using the properties of the
adjacency matrix. Finally, demonstrate how the BFS over evolving graphs can be
applied to mining citation networks.

@@

We can analyze the spread of information between the players using graph
theory. In this process, the time ordering of events matters, and hence its
graph representation $G(t) = (V(t), E(t))$ must be time dependent.
Such a graph is called an ``evolving graph''~\cite{fkp89,bkmu12},
``evolving network''~\cite{bfgm08} or ``temporal graph''~\cite{tmml09}.

Treatments of evolving graphs vary in their generality and focus.
Kivel\"a \textit{et al.}~\cite{kabg14} treat time dependence as a special
case of families of graphs with multiple interrelationships.
Others like Flajolet \textit{et al.}~\cite{fkp89} use time to index the family
of related graphs, but are not concerned with explicit time-dependent processes.
Yet others focus on incremental updates to large graphs~\cite{bkmu12}.
Here, we describe evolving graphs as a time-ordered sequence of graphs, similar
to the study of metrized graphs by Tang and coworkers~\cite{ntmm13,tmml09,tsmm09,tmml10} and of
the dynamics of communities by Grindrod, Higham and coworkers~\cite{gphe11,grihig13}.

The game described in the beginning can be encoded in an evolving graph.
The spread of information to the winner can be described in terms of traversing
this graph using discrete paths that step in both space and time.
Traversals of an ordinary (static) graph may be computed using well known methods
such as the breadth-first search (BFS). An informal description of BFS generalized
to evolving graphs can be found in \cite{tmml09}.
However, it turns out that na\"ive extensions can lead to incorrect descriptions of
the resulting graph traversals. A proper treatment requires the notion of
\emph{node activeness} to describe the set of paths that can only traverse time
or edges, which we call \textit{temporal paths}.
As a result, our treatment can be applied to any evolving graph, even those that
are highly dynamic with arbitrary changes to the nodes and edges.

In Section~\ref{sec:breadth-first-search}, we explain how the BFS algorithm can be applied correctly on
evolving graphs. Section~\ref{sec:temporal-paths} provides an example showing that considering
only products of the time-dependent adjacency matrices can lead to incorrect enumeration
of temporal paths. We present and demonstate the BFS algorithm over evolving graphs
in Section~\ref{sec:evolving-graph-bfs}, showing that it is formally equivalent to
BFS over a particular static graph generated by connecting
together active notes of the evolving graph. This static graph generates an algebraic
representation of the BFS as power iteration of its adjacency matrix to a starting
search node, as shown in Section~\ref{sec:bfsla}. The algebraic formulation also
demonstrates interesting connections between properties of the BFS algorithm and
the adjacency matrix. Finally in Section~\ref{sec:applications},
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving directed graph with 3 time stamps $t_1$, $t_2$ and $t_3$.
At each time stamp, the evolving graph is represented as a graph.
The green filled circles represent active nodes while the red circles represent
inactive nodes. Directed edges are shown as black errors.}
\label{fig:eg_shortest_path}
\end{figure}

@@

  \foreach \from/\to in {n1/n4}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=70, in=40] (n11);
  \draw[dotted,->](n10) -- (n7);
   \draw[dotted,->](n11) -- (n12);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{The two temporal paths of length 4 from $(1, t_1)$ to $(3,t_3)$ on the evolving graph
shown in Figure~\ref{fig:eg_shortest_path}, shown in black dashed lines.
The paths traverse only active nodes along edges, and are allowed to advance
between the same node if it is active at different times.}
\label{fig:active}
\end{figure}

@@

The example presented above in Section~\ref{sec:temporal-paths}
demonstrates how active nodes restrict the set of temporal paths
that need to be considered when traversing an evolving graph.

We now give a general description of the BFS algorithm over evolving graphs,
both directed and undirected, which correctly takes into account the structure
of temporal paths.
Our notation generalizes that for static graphs presented in \cite{even12,kegi11}.

@@

\begin{defn}
The \textbf{distance} from a temporal node $(v, t)$ to a temporal node $(w, s)$ is
the $k$ for which $(w, s)$ is a $k$-forward neighbor of $(v, t)$.
\end{defn}

@@

\IEEEproof
Define the set of temporal nodes
$\tilde V_L^{[t]} = \{(v_1, t) | (v_1, v_2) \in E^{[t]}\}$, which
consists of the active nodes at time $t$ which participate on the left side of an edge.
Similarly, $\tilde V_R^{[t]} = \{ (v_2, t) | (v_1, v_2) \in E^{[t]}\}$
contains the corresponding active nodes on the right side of an edge. Then
$\tilde V^{[t]} = \tilde V_L^{[t]} \cup \tilde V_R^{[t]}$ is the set of active
nodes at time $t$, and $V = \bigcup_t \tilde V^{[t]}$ is the set of all
active nodes in $G_n$.

@@

The algebraic representation of evolving graphs presented in Section \ref{sec:temp-paths-adjac}
allows us to exploit a graphical interpretation of
matrix--vector products involving the adjacency matrix~\cite{kegi11}.
If $A$ is the adjacency matrix of a (static) graph $G$ and $e_k$ is the $k$th
elementary unit vector, then the nonzero entries of $A^T e_k$ have indices that
are neighbors of $k$.
The algebraic formulation of BFS on evolving graphs follows similarly, but
requires a new kind of matrix--vector product, $\odot$, defined by
\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ can then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas nonzero entries of the other vectors represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

Consider the second iteration of BFS on $G_n$ with root $(k, t_1)$, which
requires computing the sequences

@@

The blocked matrix structure of the adjacency matrices presented here provide
interesting relationships between their matrix properties and the algorithmic
properties of BFS, made possible because of the reformulation of BFS as
repeated power iterations of the adjacency matrix in Algorithm~\ref{alg:bfsla}.
Note, however, that these matrices never need to be instantiated for practical
computations. Rather, since Algorithm~\ref{alg:bfsla} only requires the
matrix--vector product involving the adjacency matrix, the formulation of
Algorithm~\ref{alg:bfs} provides an efficient way to exploit the block structure
of $\bm A_n$. The $\odot$ operation provides an efficient way to compute the
action of the off-diagonal products. Representing the diagonal blocks
$A^{[t]}$ as sparse matrices product further reduces the cost of BFS by exploiting
latent sparsity in graphs that show up in practical applications.

\subsection{The algebraic formulation of BFS on evolving graphs}
\label{sec:bfsla}
